---
title: "Excel to Interactive Census Map"
output: html_document
Author: kimnewzealand
image : "img/portfolio/fe2013.jpg"
# showonlyimage : false
weight : 5
---


*6 April 2018*

I have been experimenting with spatial data and looking at ways to combine data objects to get different views and perspectives. This post has the following objectives:

- Import and load data direct to R without manually saving files to a local directory, so that the end to end process is reproducible.
- Use vectors to efficiently clean the Excel sheet, reducing manual steps where possible so that the process could be leveraged to clean other Excel sheets in the Census data. 
- Understand the format and structure of spatial objects and required transformations to select appropriate plotting functions.
- Join datasets to gain further insights to the original dataset.

The data used is the New Zealand 2006 and 2013 age and sex Census data and a New Zealand spatial object, with regions as the common identifiers.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages}
library(tidyverse)
# Packages for loading data
library(XLConnect)
library(readxl)
library(httr)
# Packages for data manipulation
library(stringi)
library(purrr)
# Packages for spatial files
library(rgdal) 
library(rgeos)
# Packages for geoviz
library(ggiraph)
library(ggthemes)
```

## Load New Zealand 2006 and 2013 Census data 

Firstly, get census data from the [Stats NZ website](http://archive.stats.govt.nz/Census/2013-census.aspx) using [httr](https://cran.r-project.org/web/packages/httr/index.html) and [readxl](https://cran.r-project.org/web/packages/readxl/index.html) R packages.

The 2011 Census was cancelled due to the [Christchurch earthquake](https://nzhistory.govt.nz/media/photo/census-2011).  

> The census eventually took place on 5 March 2013, two years late. The five-yearly pattern will resume when the next census is held on 6 March 2018.

```{r get Census data}
# Set the url and use GET function from httr to import the Excel file to a temporary local directory
url <- "http://archive.stats.govt.nz/~/media/Statistics/Census/2013%20Census/data-tables/regional-summary-part-1/regional-summary-tables-RC-final.xls"
GET(url, write_disk(tf <- tempfile(fileext = ".xls")))
# Read list of tables from 1st worksheet that are available in subsequent worksheets
censustables <- read_excel(tf, 1L)
# Take a look at the census tables available
head(censustables)
# Read in table 1, which is the second sheet in the Excel file
sex <- read_excel(tf,2L)
# Inspect the file layout
head(sex,20)
```
  
## Clean Excel file

Next we will clean the Census file using logical vectors and the [stringi](https://cran.r-project.org/web/packages/stringi/index.html) R package.

```{r cleanfile}
# Remove the two "Total People" columns
sex <- sex[,c(-4,-7)]
# Rename columns
names(sex) <- c("agegroup", "male2006","female2006","male2013","female2013")
# Replace the hypens with to in the agegroup column using regex and stringi
sex$agegroup <- str_replace_all(sex$agegroup, pattern = "[[:punct:]]", "to")
# Based on inspection of the file, remove header and footer file with empty space, Total regional council areas, Area Outside Region and Total New Zealand
sex <- sex[c(10:377),]
# Create a vector with the region names. From observing the file,  the regions are populated in the first column in steps of 23 and the len with is the number of rows, which should be divisible by the number of steps
steps <- 23
len <- dim(sex)[1]
region <- sex[seq(1, len, by=steps),1]
names(region) <-"region"
# Add a column with the region names filled in groups of 23 using dplyr mutate
sex <- sex %>% 
      mutate(region =rep(region$region,each=steps) )
# Create a logical vector which will be used to remove rows containing Total
t <- sex %>% select(agegroup) %>% 
            unlist() %>% 
            stri_detect_fixed('Total')
# Remove rows with Total
sex <- sex %>% 
            filter(!t)
# Use a for loop to iterate over the region$region vector and remove each region name 1 by 1. 
for (i in seq_along(region$region)) {
      # Create a logical vector r to remove rows containing region name i
      r <- sex %>% select(agegroup) %>% 
            unlist() %>% 
            stri_detect_fixed(region$region[i])
      # Remove rows with region name i
      sex <- sex %>% 
            filter(!r)
}
# Change the numerical variables from characters
sex[,c(2,3,4,5)] <- as.numeric(as.character(unlist(sex[,c(2,3,4,5)])))

```


## Load Spatial Data

Next download the spatial file with New Zealand regions REGC2017_GV_Clipped.shp to a temporary directory from the [Stats NZ website](http://archive.stats.govt.nz/browse_for_stats/Maps_and_geography/Geographic-areas/digital-boundary-files.aspx#annual).


```{r shapefile}
# Download the ESRI shapefile to the working directory, first capture the url
URL <- "http://www3.stats.govt.nz/digitalboundaries/annual/ESRI_Shapefile_2017_Digital_Boundaries_Generalised_Clipped.zip"
# Create a temporary directory
td <-  tempdir()
# Create a placeholder file
temp <- tempfile(tmpdir=td, fileext=".zip")
# Download into the placeholder file
download.file(URL, temp)
# Check the name of the REGC2017_GV_Clipped.shp file in the zip archive
unzip(temp, list=TRUE)$Name[52]
# Unzip the file to the temporary directory
unzip(temp, exdir=td, overwrite=TRUE)
# Create a new directory path for thelocation of the unzipped files as they have a subfolder
unziptd <- paste0(td,"\\",list.files(td)[1])
# Read in the REGC2017_GV_Clipped.shp files
nz_region <- readOGR(dsn = unziptd, layer="REGC2017_GV_Clipped", stringsAsFactors = F)
# unlink the temporary files and folders 
unlink(temp)
unlink(td)
unlink(unziptd)
```

## Explore Spatial Data

Let's explore the plotting options first given this spatial object following the  [introduction to spatial in R](https://cran.r-project.org/doc/contrib/intro-spatial-rl.pdf) tutorial. This nz_region file is a `r class(nz_region)[1]`. These spatial files have data and polygons "slots" which are each accessed with @. Our plotting options are initially R base plot. 

>  The generic plot() function can use Spatial* objects directly; 
>  plot is one of the most useful functions in R, as it changes its behaviour depending on the input data (this is
called polymorphism by computer scientists). Inputting another object such as plot(lnd@data) will generate
an entirely different type of plot.

Also note that the R package [ggmap](https://cran.r-project.org/web/packages/ggmap/index.html) plots objects of class ggmap and since we are using our sourced spatial object, we will not use this package in this exercise.

```{r base plot}
# Let's see a summary of the spatial object, the names refer the variables in the data slot
summary(nz_region,level=1)
# Take a look at the slot names
slotNames(nz_region)
# First plot the spatial object using base plot
plot(nz_region, col = 'light blue', border = 'blue')
# Next plot the data AREA_SQ_KM
plot(nz_region$AREA_SQ_KM)
```
  
## Transform coordinate reference system (CRS)

In order to use the the other plotting packages available we need to convert the coordinate reference system (CRS) using the spTransform function from the [rgdal](https://cran.r-project.org/web/packages/rgdal/index.html) R package as described in this [blog](https://blog.exploratory.io/making-maps-for-new-zealand-regions-and-territories-in-r-4400333df644) 
from   
`r proj4string(nz_region)`  
to  
['WGS84' (epsg:4326) is a very commonly used CRS worldwide.](https://cran.r-project.org/doc/contrib/intro-spatial-rl.pdf)

```{r transform coords}
# Transform the coordinate reference system (CRS)
wgs84 = '+proj=longlat +datum=WGS84'
nz_region_wgs <- spTransform(nz_region, CRS(wgs84))
# Confirm the coord ref is now longlat
summary(nz_region_wgs)
```
 
## Join non-spatial Census and spatial data 

Now let's join the non-spatial Census data with the data in the spatial object and create a population density variable.

```{r join data}
# Compare the region names for the Census data and the spatial data to note differences
setdiff(unique(sex$region),unique(nz_region_wgs$REGC2017_N))
# Replace the regions in the sex dataframe to match the spatial object
sex <- sex %>% 
    mutate(region = replace(region, region == "Hawketos Bay Region", "Hawke's Bay Region"), region = replace(region, region == "ManawatutoWanganui Region", "Manawatu-Wanganui Region"))
# Sum the male and female numbers grouping by region
sex_ages <- sex %>% 
      group_by(region) %>% 
      summarize_at( .vars = colnames(.)[2:5] , sum)
# Merge the nz_region and sex datasets by the region
nz_region_wgs@data <- merge(nz_region@data,sex_ages, by.x='REGC2017_N', by.y='region')
# Remove the nz_regions spatial object to save memory
rm(nz_region)
# Plot the transformed dataframe
plot(nz_region_wgs, col = 'light blue', border = 'blue')
# Remove the Area outside region which is -176 longitude to focus in this exercise on the north and south  islands.
nz_region_wgs<- subset(nz_region_wgs, REGC2017_N != "Area Outside Region")
# Plot the transformed dataframe  
plot(nz_region_wgs, col = 'blue',      border = 'white')
# Add new variables for the population density which is a relative measure than only population counts for the regions
nz_region_wgs@data <- nz_region_wgs@data %>% 
     mutate(maledens2006 = male2006/AREA_SQ_KM,
            femaledens2006 = female2006/AREA_SQ_KM,
            maledens2013 = male2013/AREA_SQ_KM,
            femaledens2013 = female2013/AREA_SQ_KM)

```
## Interactive Map ##

Next we will use ggplot2 which requires a dataframe to plot spatial objects, according to the tutorial. We will use the fortify function from the [rgeos](https://cran.r-project.org/web/packages/rgeos/index.html) R package. 


```{r fortify}
# Update the region id 'REGC2017' 
nz_region_wgs@data$id <- row.names(nz_region_wgs@data)
# Fortify the nz_region_wgs to add long and lat variables and convert to a dataframe
nz_region_f <- fortify(nz_region_wgs,id='id')
# Check the new id variable in the fortified dataframe. This refers to the number of elements in the SpatialPolygonsData
table(nz_region_f$id)
# Check the row names in nz_region_wgs 
table(nz_region_wgs$REGC2017)
# Join the new fortified and the @data
nz_region_f <- left_join(nz_region_f, nz_region_wgs@data, by="id")
# Let's check the longitude and latitude ranges to fit the map onto one page
range(nz_region_f$long)
# Now order the dataframe
nz_region_f <- nz_region_f[order(nz_region_f$order), ] 
```

Finally let's create an interactive map using [ggiraph](davidgohel.github.io/ggiraph/index.htm).

```{r interactive map}
# Compare the region names for the Census data and the transformed and fortified spatial dataframe to note differences
setdiff(unique(sex$region),unique(nz_region_wgs$REGC2017_N))
# It seems like one of the regions did not get joined. Let's temporarily fix this missing data by imputing values to get to a map prototype
nz_region_f[nz_region_f$id=="0",]$femaledens2013 <-0.6847874
nz_region_f[nz_region_f$id=="0",]$REGC2017_N <-"West Coast Region"
# Remove the nz_region_wgs spatial object to save memory
rm(nz_region_wgs)
# Create an object g with the merged dataframe with the ggplot and geom_polygon functions
nz_region_f$tooltipf <- paste0(nz_region_f$REGC2017_N, "  : ", round(nz_region_f$femaledens2013), " per sq km")
   
g <- ggplot(nz_region_f, aes(long, lat,group = group)) +
      geom_polygon_interactive(aes(fill=femaledens2013, 
                               tooltip = htmltools::htmlEscape(tooltipf, TRUE))) +
      geom_path(color = "white") + 
      coord_map() +
     scale_fill_distiller(type="seq", trans="reverse", palette = "BuPu",   name="Population density") +
      theme_void() +   
      labs(color="Pop per sq km)") +
      ggtitle("New Zealand Female Population Density in 2013")

# Now plot the g object using ggiraph
fe2013 <- ggiraph(code = print(g))
fe2013

# htmlwidgets::saveWidget(fe2013, "fe2013.html", selfcontained=FALSE) commented to out to not save separately in blogdown post
```

**Conclusions**

Loading the Census and spatial data is reproducible and the cleaning process could be leveraged for other Excel sheets. However the main challenges in this exercise were primarily in the data format transformation. Although a prototype map was produced, further investigation into the id's would needed to build an end user product. 

- Working with Spatial is slightly different to regular dataframes as some functions don't work such as the the filter from dplyr. Instead we need to use the subset function.

- Merging and joining datasets in combination with transforming data is tricky. Sanity checks and comparisons of the original data to the output is key to ensure that no data is lost in the process.

- For the interactive plots, I tried the ggplotly function from the [plotly](https://cran.r-project.org/web/packages/plotly/index.html) R package for the interactive map. It took a really long time to render and although my first choice for interactive plots is [leaflet](https://rstudio.github.io/leaflet/), I investigated alternatives and discovered ggiraph, which can be customised with CSS and used in Shiny apps.

With a prototype of the map, a Shiny App or Flexdashboard could be developed to switch between the different sexes, age groups or combined with other regional data to gain further insights.


